---
title: "SDS 439 - Homework 01"
subtitle: "Due Jan 29, 11:59pm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is an R markdown file. If you have it open in RStudio, you will see that
it is just a text file, but it has special formatting.

You will put code inside of the "code chunk" sections, like this
```{r}
x <- rnorm(7)
print(x)
```

You will complete this assignment by responding to the prompts with your
own code inside of the code chunk sections. 

You will not turn in this .Rmd file. Instead, turn in the "knitted" .pdf document,
which you can create by clicking "Knit" in the menu at the top of RStudio.

If you get an error, do not panic. Error messages contain useful information.
Stop, read what the error message says and try to follow the instructions.

You will turn in a pdf document. Your life will be easier if you can figure out
how to knit directly to a pdf. So take the time to get that to work. 

# Homework Exercises

This homework concerns analysis of the Galton families dataset, which we will use
several times throughout the course to demonstrate various models and concepts.
Each row of the dataset describes data for one child. It has columns identifying
which family the child belongs to (`family`), the father's height (`father`), the mother's height (`mother`), the number of children in the family (`children`), the ordering within the family of the present child (`childNum`), the sex of the child (`sex`), and the child's height (`childHeight`). It also has a derived variable called `midparentHeight` that
is equal to `(father + 1.08*mother)/2`.

The dataset is located in the course repository in the file `datasets/galton_families.csv`.

1. Read in the Galton Families dataset and print out the first six rows of the dataset

```{r}
```

2. Demonstrate that `midparentHeight = (father + 1.08*mother)/2`

```{r}
```

3. Make a plot of the data, showing information about the midparent height, the child   
   height, and if you can, information about the sex. Try using a different plotting symbol 
   for the two sexes. Make sure that your plot looks nice in the final pdf. This is what
   we will grade.

```{r}
```

4. Let's focus on one of the sexes, then we'll come back and analyze both of them.
   Create a new dataframe that has data for only the male children, and print out
   the first six rows.
```{r}
```

5. Use the male dataframe to perform a regression of child height on midparent height "by 
hand", which means doing all of the calculations without using the `lm` function.
Print out your estimates for the regression coefficients, their standard errors,
their t-statistics, and their p-values for the two sided test for whether the parameters equal 0. Also print out your estimate of the
residual variance.

```{r}
```

6. Why do you think that the intercept estimate has such a large standard error?
Type your answer here



7. Use the `lm` function to confirm your "by hand" calculations. Print the 
summary table from the lm fit. Use the full dataset, but tell lm to 
analyze only the males by specifying the `subset` argument of `lm`.

```{r}
```

8. Perform two more regressions using `lm` by changing the covaraiate to `mother` and
then to `father`. Print out the summary table for these two regressions.

```{r}
```

9. Use the information in the summary tables to argue which of the three models
for `childHeight` is best. Is there a biological reason why this model is best?

10. Perform a regression of `childHeight` on `midparentHeight` for the female children
and print the summary matrix.

```{r}
```

11. What are the major differences between the female model and the male model?

12. Finally, make a plot of childHeight against the midparentHeight, with the
two estimated trend lines for males and females added to the plot.

```{r}
```