---
title: "SDS 439 - Homework 02"
subtitle: "Due Feb 10, 1:00 pm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Minnesota Water Usage

This homework concerns analysis of the minnesota water dataset, which is included
in the R package for the textbook "Applied Linear Regression" by Sanford Weisberg.
The dataset has municipal and agricultural water usage data for 24 years, 
along with agricultural and municipal precipitation data for the same time period.
We are interested in studying how water usage is changing over time.

The dataset is located in the course repository in the file `datasets/mn_water.csv`.

A thorough description of the data can be found on page 32 of the `alr4` documentation:
[https://cran.r-project.org/web/packages/alr4/alr4.pdf](https://cran.r-project.org/web/packages/alr4/alr4.pdf)

1. Read in the dataset and print out the first six rows.

```{r}
water <- read.csv("../datasets/mn_water.csv")
head( water )
```

2. You should try to start your analyses by looking at the data, so you have an idea
of what to expect from your analysis. Make 4 informative plots of the data, to familiarize yourself with the major features of the data. Use good plotting practice, labeling
your axes, using legends when appropriate, etc. A good plot should show several
aspects of the data, without overwhelming you.

```{r}
plot( water$year, water$allUse, ylim = c(0, 300) )
points( water$year, water$muniUse, col = "purple" )
points( water$year, water$irrUse, col = "darkgreen" )
```

3. Use `lm` to regress the total water usage on the municipal usage and the
irrigation usage, and print out the summary table. What do the results tell
you about the relationship between these 3 variables?

```{r}
summary( lm( allUse ~ muniUse + irrUse, data = water ) )
```

4. Let's first study the municipal water usage to understand how it is
changing over time. Make a plot of municipal water usage over time.
```{r}
plot( water$year, water$muniUse, xlab = "Year", ylab = "Muni. Water Usage" )
```

5. Regress municipal water usage on year. Print the results and answer the following 
questions: How do you interpret the slope? How do you interpret the intercept?

```{r}
m1 <- lm( muniUse ~ year, data = water )
summary(m1)

# Slope is the estimated expected in crease in water usage each year
# Intercept is the estimated water usage in year 0!
```


6. To facilitate interpretation of the intercept, create a new variable 
counting the years since 1988. Then regress municipal use on your new
variable and print the results. What is the interpretation of this intercept
in this model? Verify that the intercept for this model is
equal to the fitted value for 1988 from the previous model.

```{r}
water$years_since_1988 <- water$year - 1988
summary( lm( muniUse ~ years_since_1988, data = water ) )
m1$fitted.values[24]

# Now the intercept is the estimated expect water usage in year 1988.
```

7. Perhaps the changes in municipal water usage can be explained by changes in 
municipal population. Regress municipal water use *per person* on years since
1988. BUT make sure to use an appropriate unit for the response (billions of
gallons per person is hard to interpret). Make a plot with the data and 
the regression line, and summarize your results in words.

```{r}
water$muni_gallons_per_person <- water$muniUse*1e9/water$muniPop
m1 <- lm( muni_gallons_per_person ~ years_since_1988, data = water )
plot( water$years_since_1988, water$muni_gallons_per_person )
abline(m1)
summary(m1)

# this analysis converts the response to gallons per person. We can see from
# the plot and the fit that water usage in terms of gallons per person
# is not changing much overall, though the pattern is interesting. There were
# two early years with high usage, then two later years with low usage.
# Otherwise, usage was increasing in the middle.
```

8. Let's pivot to the irrigation usage. Make two plots of the irrigation
usage, one against year, and one against agricultural precipitation

```{r}
plot( water$year, water$irrUse )
plot( water$agPrecip, water$irrUse )
```


9. Perform a multiple regression of irrigation usage on years since 1988 and 
agricultural precipitation. Perform the following steps: (1) Write out the model
in mathematical notation (like we do in class), (2) Print the summary table, (3) provide an interpretation of each of the 3 regression coefficients.

(1) $Y_i = b_0 + b_1 x_{i1} + b_2 x_{i2} + \varepsilon_i$

```{r}
# (2)
m2 <- lm( irrUse ~ agPrecip + years_since_1988, data = water )
summary(m2)
```

(3) The intercept is the expected usage in 1988 when there is 0 precipitation. 
The agPrecip coefficient is the expected change in water usage when the
precipitation increases by 1. The years_since_1988 coefficient is the
expected change in water usage each year.

10. What do the regression results tell us about how irrigation usage is 
changing over time?

It seems to be increasing over time, after controlling for precipitation.

11. Provide an estimate of how much water would have been used in 2012
if the precipitation were 12 inches. Write down the estimate in terms of
your model parameter estimates and provide the numerical value. Can you provide a standard error for your estimate?

```{r}
vv <- vcov( m2 )
est <- m2$coefficients[1] + 12*m2$coefficients[2] + (2012-1988)*m2$coefficients[3]
est
x <- c(1, 12, (2012-1988))
se <- sqrt( t(x) %*% vv %*% x )
se
```


12. Provide an estimate of how much less water would be used for irrigation
in a rainy year (say 16 inches) versus in a dry year (say 8 inches). Write down
the estimate in terms of your model parameters and provide the numerical
value. Can you provide a standard error for your estimate?

```{r}
est <- (16-8)*m2$coefficients[2]
x <- c(0, 16-8, 0)
se <- sqrt( t(x) %*% vv %*% x )
est
se
```



