---
title: "SDS 439 - Homework 04"
subtitle: "Due Mar 17, 1:00 pm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pinewood Derby

This homework concerns analysis of times from the finals of a Pinewood Derby
contest. Pinewood Derby is a competition in which people, usually children with
the help of their parents, build cars from a kit that consists of a block of 
wood and wheels. The races take place on an inclined track in a series
of heats that pit 6 cars at a time against each other.

The data are real, but the names of the children have been replaced with
fake names. You can find it in `datasets/pinewood_derby.csv`

The final was contested in order to determine which racer had the fastest
car. We are going to analyze the data with this goal in mind, using the
times as the response, and racer as a factor covariate. 

1. The first thing you should do in a data analysis is to try to find a way to 
look at the data. Read in the data and make one or two informative plots. Remember
to label your axes, use legends where appropriate, use informative plotting
symbols. Try to get multiple aspects of the data into your plot. Also, make
sure your plot looks nice when knitted to a pdf.

```{r}
```

2. To get started, create the design matrix $X$ for the one-factor model that has 
racer as the factor covariate. Create the full design matrix without dropping
any columns. Print out the first twelve rows.

```{r}
```

3. Calculate $X^T X$ and demonstrate that its inverse does not exist. This is 
actually a little tricky because the markdown will not knit if the code has
an error. You can circumvent this by putting the problematic code inside of
the `try` function, or you can use your matrix algebra knowledge to find a way
to show that the inverse does not exist (e.g. look at eigenvalues).

```{r}
```

4. Fit a model to the race times, using racer as the factor covariate. Print out
the summary of the model fit.

```{r}
```

5. Which column did R drop and how do you know?

```{r}
```

6. Using words, supported by code if necessary, answer the following questions
about the results:

(a) How do you interpret the intercept?

(b) How do you interpret the racerBess coefficient and what can you conclude
    from the t-test for the racerBess coefficient?
    
(c) What is the size of the error after accounting for racer, and how does
    it compare in magnitude to the sizes of the racer effects?
    
(d) Which racer is estimated to have the fastest car?

(e) Which racer is estimated to have the second fastest car?

(f) Do we have enough evidence to say that the fastest estimated car is 
    faster than the second-fastest? Do a t-test to answer this question.
    
```{r}
```

7. What else would you want to do in order to investigate which racers have
   the fastest cars? (You don't need to do it, just give some ideas)
   
```{r}
```